---
title: "Course Notes 15"
description: |
  An introduction to nonlinear difference equations. 
author:
  - name: Jason M. Graham 
    url: https://topicsinbiomath.netlify.app/
    affiliation: University of Scranton 
    affiliation_url: https://sites.google.com/site/jasonmgrahamus/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
creative_commons: CC BY-NC
bibliography: topics_in_biomath.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(deSolve)
library(tidyverse)
library(ggformula)
library(patchwork)
library(phaseR)
library(latex2exp)
library(DiagrammeR)
theme_set(theme_light())
```


# Goals

After reading this section of notes, you should

1) know what is a first-order nonlinear difference equation, 

2) know the procedure for determining fixed-points for first-order nonlinear difference equations, and

3) know how to use both graphical and analytical techniques for examining the stability properties of fixed-points for first-order nonlinear difference equations.

# Introduction

Let's return again to our general population model

$N(t+\Delta t) = N(t) + G(N(t),\Delta t)$

where $N(t)$ denote the number of a single isolated and homogeneous population at time $t$, and $G(N(t),\Delta t)$ is a growth function which could be positive or negative. Further, let's assume from the start that we are treating time as a discrete variable and therefore may write the model as

$N_{t+1} = N_{t} + G(N_{t})$

Previous, we took $G$ to be a linear function of $N_{t}$ and ended up with a model of the form $N_{t+1}=\lambda N_{t}$. We noted that for $N_{t+1}=\lambda N_{t}$ with $\lambda > 0$ we obtain the discrete analog of exponential growth. As in the continuous time setting, exponential growth is unrealistic. Thus, we ask the question is there a nonlinear function $G$ that may serve to "slow down"  population growth as the population approaches a carrying capacity. One option is to take

$G(N_{t}) = \rho (K - N_{t})N_{t}$

You should take a moment to think about the properties of this particular $G$ and how they relate to growth of the population. Now, with $G(N_{t}) = \rho (C - N_{t})N_{t}$ we have

$N_{t+1} = N_{t} + \rho (C - N_{t})N_{t}$

This expression may be simplified as follows:

$$
\begin{align}
N_{t+1} &= N_{t} + \rho (C - N_{t})N_{t} \\
& = N_{t} + \rho C N_{t} - \rho N_{t}^{2} \\
&= (1+\rho C) N_{t} - \rho N_{t}^{2} \\
&= (1+\rho C)\left(N_{t} - \frac{N_{t}^2}{\frac{1 + \rho C}{\rho}}\right) \\
&= (1+\rho C)N_{t}\left(1 - \frac{N_{t}}{\frac{1 + \rho C}{\rho}}\right)  \\
&= rN_{t}\left(1 - \frac{N_{t}}{K} \right)
\end{align}
$$

where $r= 1+\rho C$ and $K= \frac{1 + \rho C}{\rho}$. Thus, 

$N_{t+1}=rN_{t}\left(1 - \frac{N_{t}}{K} \right)$

is the **discrete time logistic** model. We can simplify this even further by non-dimensionalization. Let $x_{t}=\frac{N_{t}}{K}$, then

$$
\begin{align}
x_{t+1} &= \frac{N_{t+1}}{K} \\
&= \frac{1}{K}\left(rN_{t} \left(1 - \frac{N_{t}}{K} \right) \right) \\ 
&= rx_{t}(1-x_{t})
\end{align}
$$

The non-dimensional discrete logistic model $x_{t+1}=rx_{t}(1-x_{t})$ is an example of a **first-order nonlinear difference equation**. In general, first-order nonlinear difference equations take the form

$x_{t+1} = f(x_{t})$

where $f$ is a nonlinear function. Moreover, first-order nonlinear difference equations provide another class of examples of discrete time systems. 

There is already an important difference between the continuous logistic growth model

$\frac{dN}{dt}=rN\left(1 - \frac{N}{K} \right)$

and the discrete logistic growth model

$N_{t+1} = rN_{t}\left(1 - \frac{N_{t}}{K} \right)$

The continuous model non-dimensionalizes to $\frac{dx}{dt}=x(1-x)$ while the discrete model non-dimensionalizes to $x_{t+1}=rx_{t}(1-x_{t})$. In particular, we can not eliminate the parameter $r$ from the discrete time model. Thus, we must be able to assess what are the dynamics of the discrete logistic model and how are they affected by the value of $r$. We will see that $x_{t+1}=rx_{t}(1-x_{t})$ possesses much richer dynamics than its continuous time counterpart.  

# Analyzing Nonlinear Difference Equations

Let $x_{t+1}=f(x_{t})$ be a first-order nonlinear difference equation. We say that a value $\xi$ is a **fixed point** for the system if it satisfies

$\xi = f(\xi)$

Note that a fixed point $\xi$ for a difference equation $x_{t+1}=f(x_{t})$ is a constant solution to the equation in a similar way as to how an equilibrium value is a constant solution to continuous autonomous system. 

Given a difference equation $x_{t+1}=f(x_{t})$, we can find its fixed points by solving the equation

$x=f(x) \Leftrightarrow x-f(x) = 0$

for the unknown $x$. 

Graphically, fixed points for a difference equation $x_{t+1}=f(x_{t})$ correspond ot points where the line $y=x$ intersects the graph of $y=f(x)$.  

### Example

Let's determine the fixed points for the (non-dimensional) discrete logistic equation $x_{t+1}=rx_{t}(1-x_{t})$. In order to do this, we must solve

$x=rx(1-x) \Leftrightarrow x-rx(1-x) = 0 \Leftrightarrow x(1-r+rx)=0$

which has solutions $x=0$ and $x=\frac{r-1}{r}$. Note that the second fixed point is only biologically reasonable if $r \geq 1$. 

The following plot shows the intersection of $y=x$ and $y=rx(1-x)$ in the case where $r=3$. 

```{r,echo=TRUE,code_folding=TRUE}
ggplot(data=tibble(x=c(-0.1,1)),aes(x=x)) + 
    geom_function(fun=function(x) 3.0*x*(1-x)) + 
    geom_abline(slope = 1.0,intercept=0.0) + 
    geom_point(x=c(0,2/3),y=c(0,2/3),color="lightgreen",size=2) + 
    ylim(c(-0.1,1.2)) + 
    xlab("x") + 
    theme_light() + 
    theme(text = element_text(size = 18))
```

## Fixed Point Stability 

Suppose that $\xi$ is a fixed point for the difference equation $x_{t+1}=f(x_{t})$. We say that $\xi$ is **stable** if whenever $x_{0}$ is sufficiently close but not equal to $\xi$ we have that the sequence determined by $x_{t+t}=f(x_{t})$ converges to $\xi$. 

We will make use of the following fact which is analogous to linear stability analysis for continuous autonomous systems:

* If $|f'(\xi)| < 1$, then $\xi$ is stable. 

* If $|f'(\xi)| > 1$, then $\xi$ is unstable. 

* If $|f'(\xi)|=1$ then we can not draw any conclusions. 

### Example

Let's apply the stated linear stability analysis for nonlinear difference equations to the discrete logistic model $x_{t+1}=f(x_{t})=rx_{t}(1-x_{t})$. We have that $f'(x)=r-2rx_{t}$ and thus

* $|f'(0)| = r$, and

* $|f'\left(\frac{r-1}{r}\right)|=\left|2-r \right|$.

From this, we can see that $\xi=0$ is stable if $r<1$ and unstable if $r > 1$. Note that if $r<1$, then $\xi=\frac{r-1}{r}$ does not make sense biologically. 

Now, whenever $3 > r > 1$ we can see that  $\xi=\frac{r-1}{r}$ will be stable since then we will have $|2-r| < 1$. Further, if $r > 3$, then $\xi=\frac{r-1}{r}$ will be unstable. In order to know what happens when $r=1$ or $r=3$ we will need to rely on graphical techniques. 

## Cobweb Diagrams



