---
title: "Course Notes 14"
description: |
  An introduction to linear difference equations. 
author:
  - name: Jason M. Graham 
    url: https://topicsinbiomath.netlify.app/
    affiliation: University of Scranton 
    affiliation_url: https://sites.google.com/site/jasonmgrahamus/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
creative_commons: CC BY-NC
bibliography: topics_in_biomath.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(deSolve)
library(tidyverse)
library(ggformula)
library(patchwork)
library(phaseR)
library(latex2exp)
theme_set(theme_light())
```


# Goals

After reading this section of notes, you should

1) know what is a discrete-time model, 

2) know what is a linear difference equation, 

3) know how to use linear algebra techniques to analyze a linear difference equation, and

4) be aware of some biomathematical applications of linear difference equations. 

# Background

Let $N(t)$ denote the number of a single isolated and homogeneous population at time $t$. A common question is how does the population change over time? Early on, we argued that a general model for this is 

$N(t+\Delta t) = N(t) + G(N(t),\Delta t)$

which essentially states that the change (increase or decrease) in the population from $t$ to $t+\Delta t$ is determined by the current population $N(t)$ incremented by some "growth" function $G(N(t),\Delta t)$ which could be positive or negative. A first order approximation for $G$ is 

$G(N(t),\Delta t) = r \Delta t N(t)$

where we do not make any assumptions about the sign of $r$. This leads to 

$N(t+\Delta t) = N(t) + r \Delta t N(t)$

Previously, we rewrote the last expression as

$\frac{N(t+\Delta t) - N(t)}{\Delta t} = r N(t)$

and took the limit as $\Delta t \rightarrow 0$ which lead to the exponential growth model. This involves making an assumption. Specifically, we assume that time $t$ can be reasonably treated as a continuous variable. There are many situations where this an appropriate assumption. On the other hand, there are situations in which it is more reasonable to view time as a discrete variable. Let's now make this assumption in the context of our population growth model. To denote this distinction mathematically, we write $N(t)=N_{t}$, then our simple first-order population growth model becomes

$N_{t+\Delta t} = (1+r\Delta t)N_{t}$

If we increment over a fixed-size time step for $\Delta t$, then we can write the previous equation as

$N_{t+1} = \lambda N_{t}$

This is an example of a discrete time model. In general, a discrete time model is a sequence $x_{0}$, $x_{1}$, $\ldots$, $x_{n}$, $x_{n+1}$, $\ldots$, where

$x_{i}\in \mathbb{R}^{p}$ for all $i=0,1,\ldots$ and for some $p\in \mathbb{Z}^{+}$, and

$x_{i+1}=f(x_{i},x_{i-1},x_{i-2},\ldots,x_{i-k})$ for all $i=1,2,\ldots$ and some fixed $k\geq 0$.

Typically, we treat the $x_{i}$ as unknowns and refer to the expression $x_{i+1}=f(x_{i},x_{i-1},x_{i-2},\ldots,x_{i-k})$ as a **difference equation**. A **solution** of the difference equation is any sequence $x_{0},x_{1},\ldots$ that satisfies the recurrence defined by the difference equation.   

If $k=0$, then we have

$x_{i+1}=f(x_{i})$

and the discrete time model is said to be **first-order**. 

In our example model $N_{t+1} = \lambda N_{t}$ we have $f(N_{t})=\lambda N_{t}$ with $\lambda$ a constant. It is obviously a first-order model. Moreover, it is linear. In general, if a discrete time model can be written as

$x_{t+1}= A x_{t}$

where $x_{t}\in \mathbb{R}^{n}$ for all $t$ and $A$ is an $n\times n$ matrix then we say it is a **linear difference equation**. 

# Analysis of Linear Difference Equations

Suppose we have a linear difference equation

$x_{t+1}= A x_{t}$

where $x_{t}\in \mathbb{R}^{n}$ for all $t$ and $A$ is an $n\times n$ matrix, and let $x_{0}$ be some given initial vector. Then, 

$x_{1} = Ax_{0},$

$$
\begin{align}
x_{2} &= Ax_{1} \\
&= A(Ax_{0}) \\
&= A^2x_{0}
\end{align}
$$

and generally, 

$$
\begin{align}
x_{i+1} &= Ax_{i} \\
&= A(Ax_{i-1}) = A^2x_{i-1} \\
&= A(A(Ax_{i-2})) = A^3x_{i-2} \\
&= \ldots \\
&= A^{k+1}x_{i-k} \\
&= \ldots \\
&= A^{i+1}x_{0}
\end{align}
$$

Thus, the solution to a linear difference equation $x_{t+1}=Ax_{t}$ is a sequence $x_{0}$, $x_{1}=Ax_{0}$, $x_{2}=A^2x_{0}$, $\ldots$, $x_{n}=A^{n}x_{0}$, $\ldots$. Notice that to compute each term in the sequence, we must compute the $n$-fold matrix power $A^n$. Typically, this is not convenient to compute. Our goal is to show that there is another approach that can be taken. 

## Eigenvalue Analysis of Linear Difference Equations

Let $A$ be an $n\times n$ matrix and suppose that $A$ has a complete set of eigenvectors. That is, suppose that $A$ has $n$ linearly independent eigenvectors $v_{1},v_{2},\ldots, v_{n}$ with corresponding eigenvalues $\lambda_{1},\lambda_{2}\ldots,\lambda_{n}$. Now, let $x_0$ be a vector in $\mathbb{R}^{n}$. We may write 

$x_{0}=c_{1}v_{1}+c_{2}v_{2}+\cdots +c_{n}v_{n}$

where the $c_{i}$ are uniquely determined scalars. Further, if $\lambda$, $v$ is an eigenvalue-eigenvector pair for $A$, then

$A^n v = \lambda^n v,$

an identity you may derive as an exercise. Using this, we see that

$x_{i} = A^{i}x_{0}=c_{1}\lambda_{1}^i v_{1} + c_{2}\lambda_{2}^i v_{2} + \cdots + c_{n}\lambda_{n}^i v_{n}$

and this provides another way to express a solution to a linear difference equation. The benefit is that this does not require us to compute matrix powers, only powers of scalars. 

### Example

Consider the linear difference equation $x_{t+1}=Ax_{t}$ with 

$A = \left(\begin{array}{cc} 2 & 3 \\ 2 & 1  \end{array}\right),$ 

the eigenvalues of $A$ are

```{r,echo=TRUE,code_folding=TRUE}
A <- matrix(c(2,3,2,1),2,2,byrow=TRUE)
eigen(A)$values
```

One can also see that if $\lambda_{1}=4$ and $\lambda_{2}=-1$, then $v_{1}=(3,2)^{T}$ and $v_{2}=(1,-1)^{T}$ are corresponding eigenvectors. Thus, the general solution to this example linear difference equation is

$x_{i} = c_{1}(4)^{i}\left(\begin{array}{c} 3 \\2 \end{array} \right) + c_{2}(-1)^i\left(\begin{array}{c} 1 \\ -1 \end{array} \right)$

Suppose for example that $x_{0}=(1,1)^{T}$. To find the solution with this particular initial condition we must find scalars $c_{1}$ and $c_{2}$ so that 

$\left(\begin{array}{c} 1 \\1 \end{array} \right)  = c_{1}\left(\begin{array}{c} 3 \\ 2 \end{array} \right) + c_{2}\left(\begin{array}{c} 1 \\ -1 \end{array} \right)$

this can be done by solving the linear system

$\left(\begin{array}{cc} 3 & 1 \\ 2 & -1  \end{array}\right)\left(\begin{array}{c} c_{1} \\ c_{2} \end{array} \right) = \left(\begin{array}{c} 1 \\1 \end{array} \right)$

which is

```{r,echo=TRUE,code_folding=TRUE}
V <- matrix(c(3,1,2,-1),2,2,byrow = TRUE)
b <- c(1,1)
(x<-solve(V,b))
```


So, the solution to $x_{t+1}=Ax_{t}$ with 

$A = \left(\begin{array}{cc} 2 & 3 \\ 2 & 1  \end{array}\right),$

and $x_{0}=(1,1)^{T}$  is

$x_{t}=\frac{2}{5}(4)^t \left(\begin{array}{c} 3 \\ 2 \end{array} \right) - \frac{1}{5}(-1)^t \left(\begin{array}{c} 1 \\ -1 \end{array} \right) = (4)^t \left(\begin{array}{c} \frac{6}{5} \\ \frac{4}{5} \end{array} \right) + (-1)^t \left(\begin{array}{c} -\frac{1}{5} \\ \frac{1}{5} \end{array} \right)$

An interesting question about the dynamics of discrete time models is, what happens to the system as $t\rightarrow \infty$. Note that in the expression

$x_{i} = c_{1}\lambda_{1}^i v_{1} + c_{2}\lambda_{2}^i v_{2} + \cdots + c_{n}\lambda_{n}^i v_{n}$

if all of the eigenvalues satisfy $|\lambda_{i}|<1$, then $x_{i} \rightarrow 0$ as $i \rightarrow \infty$. Thus, if $x_{t+1}=Ax_{t}$ is a linear difference equation, we expect $x_{t} \rightarrow 0$ in the long-term whenever all of the eigenvalues of $A$ have absolute value less than 1. 

We can apply the result we just derived to our simple population growth model $N_{t+1}=\lambda N_{t}$ by observing that it is a one-dimensional linear difference equation. Therefore, the general solution is $N_{t}=c\lambda^t N_{0}$ and then the population will die out if $|\lambda| < 1$. On the other hand, what will happen if $|\lambda| > 1$? In the one-dimensional case it is pretty clear, the magnitude of the population will grow without bound. What happens in the general case $x_{t+1}=Ax_{t}$ whenever $A$ is an $n\times n$ matrix? 

To answer this question, suppose that our eigenvalues are ordered so that $|\lambda_1| > |\lambda_2| > \cdots > |\lambda_{n}|$ and then rewrite

$$
\begin{align}
x_{t} &= c_{1}(\lambda_{1})^tv_{1} + c_{2}(\lambda_{2})^tv_{2} + \cdots + c_{n}(\lambda_{n})^tv_{n} \\
&= \lambda_{1}^t\left(c_{1} v_{1} + c_{2}\left(\frac{\lambda_{2}}{\lambda_{1}}\right)^t v_{2} + \cdots + c_{n}\left( \frac{\lambda_{n}}{\lambda_{1}}\right)^t v_{n}\right)
\end{align}
$$

Now, for $j=2,\ldots n$ we have that $\left| \frac{\lambda_{j}}{\lambda_{1}}\right| < 1$. So, we expect that as $t\rightarrow \infty$ each of these terms will go to zero and $x_{t}$ will asymptotically approach the line determined by the vector $v_{1}$. For a matrix $A$, we call the eigenvalue of $A$ that is largest in absolute value the **dominant eigenvalue**. Thus, the dominant eigenvalue and corresponding eigenvector determine the long-term dynamics of a discrete dynamical system determined by a linear difference equation.  

### Example

Consider again the linear difference equation $x_{t+1}=Ax_{t}$ with 

$A = \left(\begin{array}{cc} 2 & 3 \\ 2 & 1  \end{array}\right)$

This matrix has dominant eigenvalue $\lambda_{1}=4$ with corresponding eigenvector $v_{1}=(3,2)^{T}$. The following figure shows the points $x_{t}$ in the plane determined by the sequence $x_{t}=A^{t}x_{0}$ together with the line through the vector $v_{1}$.

```{r}
N <- 50
x <- numeric(N)
y <- numeric(N)
z <- numeric(N)
x[1] <- 1
y[1] <- 1
z[1] <- 2/3 * x[1]
A <- matrix(c(2,3,2,1),2,2,byrow = TRUE)
for (i in 2:N){
  x[i] <- sum(A[1, ]*c(x[i-1],y[i-1]))
  y[i] <- sum(A[2, ]*c(x[i-1],y[i-1]))
  z[i] <- 2/3 * x[i]
}

lde_df <- tibble(x=x,y=y,z=z)
lde_df %>% ggplot() + geom_point(aes(x=log10(x),y=log10(y))) + 
  geom_line(aes(x=log10(x),y=log10(z)))
```

Note that we have plotted on the log scale because there is exponential growth in the magnitude of points in the sequence of vectors $x_{t}$. This plot clearly shows how the asymptotic dynamics of the discrete time model "follows" the direction of the eigenvector corresponding with the dominant eigenvalue. 


