---
title: "R for Biomathematics"
description: |
  An Introduction to R for Biomathematics
author:
  - name: Jason M. Graham 
    url: https://topicsinbiomath.netlify.app/
    affiliation: University of Scranton 
    affiliation_url: https://sites.google.com/site/jasonmgrahamus/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
creative_commons: CC BY-NC
bibliography: topics_in_biomath.bib
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggformula)
library(deSolve)
library(phaseR)
```

# Computation in Biomathematics

As in other scientific disciplines, numerical and scientific computing play a central role in 
biomathematics. For the purposes of *Topics in Biomathematics*, we are mainly interested in the numerical solution of mathematical problems, *e.g.,* ordinary differential equations, and scientific visualization, *e.g.,* plotting functions. Numerous computing platforms or environments are used in biomathematics. Some notable options include

* [MATLAB](https://www.mathworks.com/products/matlab.html) or [Octave](https://www.gnu.org/software/octave/index), 

* [Maple](https://www.maplesoft.com/), 

* [Python](https://www.python.org/), 

* [XPPAUT](http://www.math.pitt.edu/~bard/xpp/xpp.html), 

* [R](https://www.r-project.org/), and

* [Julia](https://julialang.org/). 


The purpose of this document is to provide an introduction to parts of the R computing ecosystem that are particularly useful for biomathematics. 

# What is R? 

[R](https://en.wikipedia.org/wiki/R_(programming_language)) is an open-source statistical computing platform that was developed as an alternative to proprietary statistical software. In the years since it's initial release, R has become immensely popular and a vast ecosystem of open-source software has developed around R that greatly enhances it's utility in science.   

# Why R?


Given many other options, why do we focus on R as the computing tool of choice in *Topics in Biomathematics*? The principal reasons are:

1) R is free but also well-developed and widely-supported. 

2) Even though it was developed for statistical applications, R works well for numerical computing of the sort that is useful in biomathematics. 

3) The [RStudio](https://www.rstudio.com/) integrated development environment (IDE) facilitates not only coding in R but also presenting and sharing results so as to align nicely with the course objective "Communicate effectively in writing their work in applying the techniques of applied mathematics to problems arising in the life sciences.". 

4) Many biologists "speak" R and it is very helpful to have a common language when collaborating on interdisciplinary problems. 


# An Introduction to R

In what follows, we illustrate some R commands that will be used commonly throughout the *Topics in Biomathematics* course. This document is by no means comprehensive. The best ways to learn more about R are

1) by using the [`swirl`](https://swirlstats.com/) package, and 

2) searching the web for specific topics. 

Here is a short video about `swirl`:

```{r swirl_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/olOJfAjzdO8") %>%
  vembedr::use_align("center")
```


## Installing R and RStudio

R and RStudio are both free and easy to install. The videos below can serve as a helpful guide if you want to use them. Alternatively, you can access R and RStudio through a web browser without direct installation via [RStudio Cloud](https://rstudio.cloud/). This does require an account but is free to set up.

### Installation on Mac

Read [this article](https://macpaw.com/how-to/download-r-on-mac?campaign=cmmx_search_dsa_us_en&ci=10147083349&adgroupid=123618298718&adpos=&ck=&targetid=dsa-1280095891351&match=&gnetwork=s&creative=523598858635&placement=&placecat=&accname=cmm&gclid=CjwKCAiA55mPBhBOEiwANmzoQjy_N3sWjtX395AVPNBDvJVDsd2mKXLboj3XikILm3OcCx7edQqcVhoC5d8QAvD_BwE) or watch this video:

```{r mac_install,echo=FALSE}
vembedr::embed_url("https://youtu.be/AEebOXiMyyI") %>%
  vembedr::use_align("center")
```


### Installation on PC

Read [this article](https://medium.com/@GalarnykMichael/install-r-and-rstudio-on-windows-5f503f708027) or watch this video:

```{r pc_install,echo=FALSE}
vembedr::embed_url("https://youtu.be/GAGUDL-4aVw") %>%
  vembedr::use_align("center")
```

### RStudio Cloud

Here is an intro to RStudio Cloud:

```{r rstudio_cloud,echo=FALSE}
vembedr::embed_url("https://youtu.be/uK1Va_UWQFc") %>%
  vembedr::use_align("center")
```

## R as a Calculator

R can be used for basic arithmetic calculations:

```{r arithmetic}
(1 + 2) # addition
(6 - 4) # subtraction 
(3*5) # multiplication 
(8/3) # division
(2^3) # exponentiation
```

There are also a number of mathematical functions:

```{r math_functions}
(exp(1))
(log(1))
(log(exp(2)))
(log10(10))
(sin(pi/4))
(cos(pi/3))
```

Sometimes it is convenient to store a value as a variable. The follow code stores the value of 3 as a variable `x` and then multiplies it by 1.5:

```{r a_variable}
x <- 3
1.5*x
```

In R, the notation `<-` is the assignment operator.  

You can do arithmetic with variables just like you can with numbers. For example, 

```{r variable_arihtmetic}
x <- 3
y <- 10
(x + y)
(x*y)
(y^x)
```

You should always pay close attention to the order of operations when doing arithmetic in R. Use parentheses to make sure you compute what you mean to compute. For example, 

```{r op_order_1}
((3 - 2) * 6)
```

versus

```{r op_order_2}
(3 - (2 * 6))
```


## Data in R

There are several common ways to represent different types of data in R. In this section we cover those that are most relevant to biomathematics. 

### Numeric Vectors

In R, we create a vector or array of values using concatenation with the operator `c`. For example, 

```{r a_vector}
(a <- c(1,5,-7))
```

creates a numeric vector with entries 1, 5, and -7, then we store the result as a variable called `a`. 

We can perform arithmetic operations on vectors. For example, 

```{r vector_arithmetic}
b <- c(2,-1,6)
(a + b) # vector addition
(a-b) # vector subtraction
(a*b) # elementwise vector multiplication 
(a^b) # elementwise vector exponentiation 
```

**Question:** In the last code block, what is happening to produce the result corresponding to `a^b`? 

We can also call math functions on vectors. For example, 

```{r vector_math_funcs}
x <- c(1,2,3)
(exp(x))
(cos(pi*x))
```

Note that these math functions are "vectorized" in that the operate elementwise on a vector. 

### Character Vectors

We can also create so-called character vectors in R. For example, 

```{r char_vect}
(A <- c("calculus","linear algebra", "differential equations"))
```

We won't make extensive use of character vectors in *Topics in Biomathematics* but it is useful to know that they exist. 

### Matrices in R

The `matrix` functions allows us to create a matrix in R. For example, 

```{r r_matrix}
(M <- matrix(c(1,2,3,4),2,2))
(N <- matrix(c(1,2,3,4,5,6),2,3))
(P <- matrix(c(1,2,3,4,5,6),3,2))
(Q <- matrix(c(1,2,3,4,5,6),2,3,byrow=TRUE))
```

### Lists in R

All of the entries in a vector or a matrix must by of the same type. For example, all numeric values or all characters. In R, lists are a data structure that allow us to store data of mixed types. For example, 

```{r r_lists}
(l1 <- list(c(1,2,3),c("jason","graham"),matrix(c(1,2,3,4),2,2)))
(l2 <- list(initial=c(1,12),times=seq(0,10,by=1),values=matrix(1:22,11,2))) # list with names
```

Note that the entries of a list can be accessed using either indexing or names (provided that the list entries are named). For example,

```{r list_accessing}
(l1[[2]])
(l2[[1]])
(l2$initial)
```

If a list has named elements, we can access the names using the `names` function. For example, 

```{r list_names}
names(l2)
```

Here is a helpful video on R lists:

```{r list_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/Px9VNWHja4M") %>%
  vembedr::use_align("center")
```

### Data Frames in R

A data frame in R is a data structure that corresponds to something like a spreadsheet. A data frame is a special type of list object. You can think of a data frame as a cross between a list and a matrix. Here is how you ceate a data frame:

```{r data_frame}
(my_data <- data.frame(course=c("MATH 114","MATH 221","MATH 222","MATH 351", "MATH 341"),
                       num_students=c(10,12,8,5,7),
                       level=c("F","F","S","S","T")))
```

Note that a data frame contains rows and columns, every column is a vector (either numeric or character), and every column has the same number of entries. The columns of a data frame all have names. We can also use the `names` function to determine the column names of a data frame. For example, 

```{r data_frame_names}
names(my_data)
```

Typically we do not create a data frame by hand, but data is often read in as a data frame. The principal use of data frames in R is that they are the best way to provide data to plotting functions. 

Here is a helpful video on R data frames:

```{r df_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/9f2g7RN5N0I") %>%
  vembedr::use_align("center")
```

## Linear Algebra in R

Linear algebra plays a central role in solving problem in biomathematics. This section demonstrates some linear algebra computations that will be used frequently in *Topics in Biomathematics* performed in R. 

### Matrix Arithmetic

If $A$ and $B$ are matrices of the same size then we can add them. Matrix addition is computed elementwise. We demonstrate this in R.

```{r r_matrix_add}
(A <- matrix(c(1,2,-1,3,1,0,2,-1,2),3,3,byrow = TRUE))
(B <- matrix(c(3,2,-1,2,1,-1,2,-3,1),3,3,byrow = TRUE))
A + B # matrix addition
```


We can also multiply a matrix by a number (scalar multiplication). In R

```{r r_scalar_mult}
3 * A
```

Recall that if $A$ is an $m\times n$ matrix and $B$ is an $n \times p$ matrix, then we can form the [matrix product](https://en.wikipedia.org/wiki/Matrix_multiplication) $AB$ which will be a matrix of size $m\times p$. We demonstrate this in R.

```{r r_matrix_mult}
(A <- matrix(c(1,2,-1,3,1,0,2,-1,2),3,3,byrow = TRUE))
(B <- matrix(c(3,2,-1,2,1,-1),3,2,byrow = TRUE))
A %*% B # matrix multiplication
```

A special case of matrix multiplication is matrix-vector multiplication. For example, 

```{r mat_vect_mult}
(v <- c(1,-1,1))
A %*% v # matrix-vector multiplication 
```

### Matrix Transpose and Determinant

In R, the transpose of a matrix is obtained as follows:

```{r r_transpose}
t(B)
```

and the determinant of a square matrix is computed as

```{r r_det}
det(A)
```


### Eigenvalues and Eigenvectors

Recall that if $A$ is a square matrix ($n\times n$), then an eigenvalue-eigenvector pair is a number $\lambda$ and a non-zero vector $v$ satisfying $Av=\lambda v$. In R, we can compute the eigenvalues and eigenvectors for a square matrix as

```{r}
(evals_evects <- eigen(A))
```

Note that the result is an R list that consists of 

1) a vector that contains the eigenvalues, and 

2) a matrix, the columns of which are the eigenvectors.

It is a fact from linear algebra that the determinant of a square matrix is the product of its eigenvalues. Let's confirm this with our example matrix:

```{r evals_det}
(det(A))
(prod(evals_evects$values))
```

Let's also confirm that $Av=\lambda v$ for the first eigenvalue-eigenvector pair we just computed. 

```{r check_eigens}
(A %*% evals_evects$vectors[ ,1])
(evals_evects$values[1] * evals_evects$vectors[ ,1])
```

## Functions in R

A function is a programming construct that takes some number of inputs (possibly zero), performs some operations, and returns a value (which may be null). Writing functions is useful in general because it saves from having to rewrite the same code over and over again, makes code easier to read, and facilitates reproducibility. Additionally, code to compute numerical solutions to differential equations requires us to write a function specifying the differential equation. Thus, we are motivated to learn a little about writing functions in R. 
Suppose we want to create a function that inputs a number, multiplies that number by 2, and returns the result. Here is how you do this:

```{r mult_by_two_func}
mult_by_two <- function(x){
  return(2*x)
}
```

Let's call our function:

```{r call_mult_by_two_func}
mult_by_two(3)
```

We can also do the following:

```{r alt_call_mult_by_two_func}
x <- 3
mult_by_two(x)
```

What if we want to multiply by an arbitrary number? One way to do this is as follows:

```{r mult_by_a}
mult_by_a <- function(x,a=2){
  return(a*x)
}
```

In the function `mult_by_a` we have an optional argument with a default value. If `a` is not specified, we multiply by 2, if a value for `a` is given then it is used. For example:

```{r call_mult_by_a}
(mult_by_a(3))
(mult_by_a(3,a=4)) # a is given by name
x <- 4
a <- 6
(mult_by_a(x))
(mult_by_a(x,a)) # a is given by position 
```

Let's try something a little more interesting. Suppose we want to compute the value of an arbitrary polynomial function. Since the coefficients uniquely determine the polynomial and since R can perform vectorized operations, we can do the following:

```{r poly_func}
poly_func <- function(x,coeff_vect=c(1)){
  degree <- length(coeff_vect) - 1 
  powers <- 0:degree
  x_powers <- x^powers
  return(sum(coeff_vect*x_powers))
}
```

Let's test our function:

```{r test_poly_func}
(poly_func(2))
(poly_func(2,coeff_vect=c(3)))
(poly_func(2,coeff_vect=c(0,1)))
(poly_func(2,coeff_vect=c(0,0,1)))
(poly_func(2,coeff_vect=c(1,2,3)))
```

Later we will see more examples of functions in R. In you want to learn more before then, here is a helpful video on writing functions in R:

```{r functions_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/p8tAQx7ijXE") %>%
  vembedr::use_align("center")
```

## R Packages

R packages extend the functionality of R by implementing functions for various tasks. Packages must be installed before they can be used. Installing packages is easy to do. Either use the `install.packages` function or in RStudio use the Install icon on the Packages tab. 

Once you have installed the package or packages you want to use, you can load them with the `library` function. For *Topics in Biomathematics* we will make heavy use of the plotting packages [`ggplot2`](https://ggplot2.tidyverse.org/) and [`ggformula`](http://www.mosaic-web.org/ggformula/), and the packages [`deSolve`](https://desolve.r-forge.r-project.org/) and [`phaseR`](https://www.rdocumentation.org/packages/phaseR/versions/2.1.3) for working with differential equations. To install these packages one can use

```{r install_packages,eval=FALSE}
install.packages(c("ggplot2","ggformula","deSolve","phaseR"))
```

You only ever have to install a package once. To load these packages, you run

```{r load_packages,eval=FALSE}
library(ggplot2)
library(ggformula)
library(deSolve)
library(phaseR)
```

Packages must be loaded every time you begin a new R session.  

## Visualization in R


When you present your work you should use figures and plots to clearly convey your results.  As mentioned, the two packages [`ggplot2`](https://ggplot2.tidyverse.org/) and [`ggformula`](http://www.mosaic-web.org/ggformula/) make it possible to create high-quality plots in R. These packages are powerful and extensive so we will not be able to cover their full functionality here. Hopefully what is included here is enough to get you started, at least for the purposes of *Topics in Biomathematics*. It should be noted that `ggformula` is built on top of `ggplot2`. 

### Plotting a Function

Here are two ways to plot a simple mathematical function of the form $y=f(x)$ in R:

Suppose we want to plot $f(x) = x\cos(x)$ over $[0,4\pi]$. Then we can do

```{r function_plot_ggformula}
gf_function(fun = function(x){x*cos(x)},xlim=c(0,4*pi))
```

or

```{r function_plot_ggplot2}
ggplot(data.frame(x=c(0,4*pi)),aes(x=x)) + 
  geom_function(fun = function(x){x*cos(x)})
```

We can modify our plots in various ways. For example, 

```{r function_plot_ggformula_mod}
gf_function(fun = function(x){x*cos(x)},xlim=c(0,4*pi),color="blue",lwd=2)
```

or

```{r function_plot_ggplot2_mod}
ggplot(data.frame(x=c(0,4*pi)),aes(x=x)) + 
  geom_function(fun = function(x){x*cos(x)},color="blue",lwd=2)
```

### More on Plots

It is possible to create a variety of different types of plots using [`ggplot2`](https://ggplot2.tidyverse.org/) and [`ggformula`](http://www.mosaic-web.org/ggformula/), and plots creating using these packages can be highly customized. As we go through the *Topics in Biomathematics* course we will learn a little more about the utility of these packages. For now, we refer anyone interested to learn more to this helpful video on `ggplot2`:

```{r ggplot_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/iOosTTQwCW4") %>%
  vembedr::use_align("center")
```

[This article](https://stat2labs.sites.grinnell.edu/Handouts/rtutorials/IntroToggformula.html) provides more information on `ggformula`. 

## Numerical Solution of Differential Equations in R

A problem that we will often encounter is to obtain a numerical solution to an initial value problem (IVP) for an ordinary differential equation (ODE) or a system of ordinary differential equations (ODEs). An IVP for an ODE or ODEs typically takes the form

$\frac{d{\bf x}}{dt} = {\bf F}({\bf x}(t),t)$,

${\bf x}(t_{0}) = {\bf x}_{0}$.

Here ${\bf x}$ is the unknown function of time and ${\bf F}$ is the right hand side function. These may be scalars (in case of a single ODE) or vectors (in case of a system of ODEs). First, we will focus on the case of solving a single scalar ODE, then we will examine systems on ODEs. 

For example, we might want to obtain a numerical solution to 

$\frac{dN}{dt} = r N$,

$N(0) = 10$,

with $r=0.5$ and over the time interval $[0, 5]$.

In order to obtain a numerical solution to an IVP for ODEs we must provide at minimum the following information to the computer:

  1) A function that describes the right hand side ${\bf F}({\bf x}(t),t)$. Note that ${\bf F}$ may involve parameters whose values will also need to be specified. 
  
  2) A time interval, $[t_{0},t_{f}]$, over which to obtain a numerical solution.
  
  3) Initial conditions ${\bf x}_{0}$. 

Finally, once we have obtained numerical solutions to an IVP for ODEs it is typically desirable to plot these solutions in one way or another. We will also see examples of ways to plot our numerical solutions in R. 

### Solving a Single ODE

To use the methods described in this section you must make sure that the `deSolve` package is installed and loaded. The main function for obtaining numerical solutions in the `deSolve` package is the `ode` function. In order to use it, we must write a function in R syntax that describes the right hand side of our differential equation. Let's do this for our example problem

$\frac{dN}{dt} = r N$,

$N(0) = 10$,

with $r=0.5$ and over the time interval $[0, 5]$.

```{r single_ode}
My_RHS <- function(t,state,parameters){
  with(as.list(c(state,parameters)),{
    dN <- r*N
    
    list(c(dN))
  })
}
```

In addition to describing the right hand side of the differential equation, we must also specify the times at which to obtain the numerical solution and the initial condition. This is done as follows:

```{r ode_data}
state <- c(N=10)
times <- seq(0,5,by=0.1)
```

Now we are ready to obtain the numerical solution:

```{r solve_example}
solN <- ode(y=state,times=times,func=My_RHS,parms=c(r=0.5))
```

Let's look at what has been returned by our call to the numerical solver, the results of which are stored in the variable `solN`. 

```{r print_solution}
solN
```

The method from the `deSolve` package returns a **matrix** where the first column contains the time values at which we obtained numerical solution values and another column that contains the solution values corresponding to each of the time points. We can easily plot the solution values versus the time points as follows:

```{r plot_example}
plot(solN,col="blue")
```


A nicer plot can be obtained as follows:

```{r plot_example_ggplot}
ggplot(as.data.frame(solN),aes(x=time,y=N)) + geom_point()
```

If we want a curve instead:

```{r plot_example_ggplot_line}
ggplot(as.data.frame(solN),aes(x=time,y=N)) + geom_line(lwd=1)
```

We can plot the numerical solution together with the analytical solution $N(t) = 10e^{0.5 t}$ as follows:

```{r example_analtiycal}
ggplot(as.data.frame(solN),aes(x=time,y=N)) + geom_point() + 
  geom_function(fun=function(t) 10*exp(0.5*t),color="blue")
```

**Exercise:** Modify what we have done using the `deSolve` package to obtain and plot a numerical solution to the IVP 

$\frac{dN}{dt} = r N$,

$N(0) = 5$,

with $r=2.5$ and over the time interval $[0, 10]$.

### Solving a System of ODEs

Consider the following problem:

$\frac{dx}{dt}=-R_{0}xy$

$\frac{dy}{dt}=R_{0}xy - y$

with $R_{0} = 0.2$ and $x_{0} = 0.95$ and $y_{0} = 0.05$.

We can code this system of ODEs in R as:

```{r system_example}
My_RHS <- function(t,state,parameters){
  with(as.list(c(state,parameters)),{
    dx <- -R0*x*y
    dy <- R0*x*y - y
    
    list(c(dx,dy))
  })
}
```

We must specify our time interval and initial conditions:

```{r system_data}
state <- c(x=0.95,y=0.05)
times <- seq(0,20,by=0.1)
```

Now, we are ready to obtain the numerical solution:

```{r system_sol}
SIR_sol <- ode(y=state,times=times,func=My_RHS,parms=c(R0=0.2))
```

Again, the method from the `deSolve` package returns a matrix where the first column contains the time values at which we obtained numerical solution values and another column that contains the solution values corresponding to each of the time points. We can easily plot the solution values versus the time points as follows:

```{r plot_system}
plot(SIR_sol,col="blue")
```

**Question:** Again, what happens if we change the value of $R_{0}$ to $R_{0}=1.5$?

Often we want to have greater control over our plots and to be able to make high quality figures. The best way to do this in R is to use the capabilities of the `ggplot2` package. In order to do this, we will first convert the matrix of solutions returned by the `ode` function from the `deSolve` package into a data frame. This is basically like an spread sheet format for data. 

```{r system_df}
SIR_sol_df <- as.data.frame(SIR_sol)
```

We can get a sense of the structure of this data by viewing the first few rows:

```{r system_df_head}
head(SIR_sol_df)
```

It is convenient to change the names of the columns to something more meaningful as follows:

```{r rename_system_df}
names(SIR_sol_df) <- c("time","susceptible","infected")
```

Let's look at the first few rows again:

```{r system_df_head_again}
head(SIR_sol_df)
```

Now, we can obtain plots by doing commands like:

```{r system_plot_b}
ggplot(data=SIR_sol_df,mapping=aes(x=time,y=susceptible)) + geom_line()
```

or

```{r system_plot_c}
ggplot(data=SIR_sol_df,mapping=aes(x=time,y=infected)) + geom_line()
```

Here are some other plotting alternatives:

```{r system_plot_d}
 SIR_df_long <- pivot_longer(SIR_sol_df,-time,names_to = "compartment",values_to = "population")
ggplot(SIR_df_long,aes(x=time,y=population,color=compartment)) + geom_line(lwd=1)
```

or 

```{r system_plot_e}
ggplot(SIR_df_long,aes(x=time,y=population)) + geom_line(lwd=1) + facet_wrap(~compartment)
```

We can also plot $x(t)$ and $y(t)$ together as parametric curves:

```{r system_plot_f}
ggplot(SIR_sol_df,aes(x=susceptible,y=infected)) + geom_path(lwd=1)
```

Using color to indicate increasing values of time can be interesting:

```{r system_plot_g}
ggplot(SIR_sol_df,aes(x=susceptible,y=infected,color=time)) + geom_path(lwd=1)
```

This color scale may not be the best, let's change it:

```{r system_plot_h}
ggplot(SIR_sol_df,aes(x=susceptible,y=infected,color=time)) + 
  geom_path(lwd=1) +
  scale_color_viridis_c()
```

The book [@soetaertSolvingDEsR2012] covers much more on numerical computing with differential equations in R.  

## Phase-Plane Methods in R

The `phaseR` package facilitates the mix of qualitative and quantitative methods for the analysis of 1- and 2-dimensional autonomous systems of ODEs. We illustrate its usage. 

### PhaseR for Scalar Equations

A first-order autonmous scalar ordinary differential equation is one with the general form

$\frac{dx}{dt} = f(x)$,

where $f(x)$ is a continuous (usually even differentiable) function that does not depend explicitly on time $t$ (this is the reason for the word autonomous). Most often, we are interested in the case when $f(x)$ is a nonlinear function and we sometimes refer to $\frac{dx}{dt} = f(x)$ as a one-dimensional system. 

**Example**

Consider as an example the equation

$\frac{dx}{dt} = x(1-x)$. 

The left hand side is the derivative $\frac{dx}{dt}$ and hence the (local) rate of change of $x$ as a function of time $t$. Therefore, because of the equal sign, $f(x)$ must also provide information about the (local) rate of change of $x$ but now for various "state" values, that is, values of $x$. 

We can use `phaseR` to analyze this problem. 

First we need to describe the differential equation:

```{r phaseR_scalar}
nondim_logistic <- function (t, x, parameters) 
{
    list(x * (1 - x))
}
```

Now we plot what is called the phase portrait:

```{r phase_line}
nondim_logistic_phasePortrait <- phasePortrait(nondim_logistic,
                                        ylim   = c(-0.5, 1.5),
                                        points = 10,
                                        frac   = 0.5)
```

The arrows here tell us, starting from an initial state value $x$, whether the solution to the differential equation will increase as a function of time or decrease as a function of time. 

Further, we can use `phaseR` to obtain some "trajectories" to our ODE:

```{r scalar_trajectories,message=FALSE,warning=FALSE}
nondim_logistic_flow <- flowField(nondim_logistic,
                                 xlim       = c(0, 10),
                                 ylim       = c(-0.5, 1.5),
                                 parameters = NULL,
                                 points     = 15,
                                 system     = "one.dim",add=FALSE)
nondim_logistic_solutions <- trajectory(nondim_logistic,
                                        y0   = c(0,0.5,1,1.5),
                                        tlim=c(0,10),
                                        system = "one.dim")
```

There is an general function in `phaseR` that will carry out a complete qualitative analysis for an autonomous system (either one-dimensional or two-dimensional). For example, call the following function in the R console:

```{r ,eval=FALSE}
phasePlaneAnalysis(nondim_logistic,xlim = c(0,10),ylim = c(-1.5,1.5),system = "one.dim")
```


### PhaseR for Two-Dimensional Systems

Let's use `phaseR` to explore the dynamics of two-dimensional linear autonomous systems. 

Consider for example: 

$\frac{d{\bf x}}{dt} = A {\bf x}$, 

where $A$ is a $2\times 2$ matrix of the form

$A = \left(\begin{array}{cc} a & 0 \\ 0 & -1 \end{array}\right)$.

Let's examine how various values of $a$ change the dynamics (specifically the phase portraits) of the systems. Observe that as long as $a\neq 0$ there is a unique steady-state ${\bf x}^{\ast}={\bf 0}$. 

#### Phase Portraits for Varying $a$

```{r linear_system}
simple_linear <- function(t,state,parameters){
  with(as.list(c(state,parameters)),{
    dx <- a*state[1]
    dy <- -state[2]
    
    list(c(dx,dy))
  })
}
```


```{r examp_parm}
a_val <- -2
```


```{r system_pp_a}
simpleLinear_flowField  <- flowField(simple_linear,
                                       xlim       = c(-2, 2),
                                       ylim       = c(-2, 2),
                                       parameters = c(a=a_val),
                                       points     = 19,
                                       add = FALSE)
```

Let's add some solution curves:

```{r system_pp_b}
simpleLinear_flowField  <- flowField(simple_linear,
                                       xlim       = c(-2, 2),
                                       ylim       = c(-2, 2),
                                       parameters = c(a=a_val),
                                       points     = 19,
                                       add = FALSE)
state                        <- matrix(c(1,1,1,-1,-1,1,-1,-1,0,2,0,-2,-1,0,1,0),
                                    8, 2, byrow = TRUE)
simpleLinear_trajectory <- trajectory(simple_linear,
                                        y0         = state,
                                        tlim       = c(0, 10),
                                        parameters = c(a=a_val),add=TRUE)
```

Let's plot the numerical solutions as a function of time for specific initial conditions:

```{r linear_system_sol}
state <- c(x=1,y=1)
times <- seq(0,10,by=0.05)
simple_linear_sol <- ode(y=state,times=times,func=simple_linear,parms=c(a=a_val))
plot(simple_linear_sol,col="blue")
```

#### Two-dimensional Non-Linear Autonomous Systems

Generally, a two-dimensional autonomous system has the form

$$
\begin{align*}
\frac{dx}{dt} &= f(x,y) \\
\frac{dy}{dt} &= g(x,y), 
\end{align*}
$$
Such as system is sometimes called a vector field. 

Let's adapt our code to examine some phase portraits for a nonlinear system. 

```{r nonlinear_pp_a}
non_linear <- function(t,state,parameters){
  with(as.list(c(state,parameters)),{
    x <- state[1]
    y <- state[2]
    dx <- -x + x^3
    dy <- -2*y
    
    list(c(dx,dy))
  })
}
```

First we plot the vector field:

```{r nonlinear_pp_b}
nonLinear_flowField  <- flowField(non_linear,
                                       xlim       = c(-2, 2),
                                       ylim       = c(-2, 2),
                                       parameters = NULL,
                                       points     = 19,
                                       add = FALSE)
```

Now we can obtain a more detailed picture of the phase portrait:

```{r nonlinear_pp_c}
nonLinear_flowField  <- flowField(non_linear,
                                       xlim       = c(-2, 2),
                                       ylim       = c(-2, 2),
                                       parameters = NULL,
                                       points     = 19,
                                       add = FALSE)
nonLinear_nullclines <- nullclines(non_linear,
                                        xlim       = c(-2, 2),
                                        ylim       = c(-2, 2),
                                        parameters = NULL,
                                        points     = 500)
state                        <- matrix(c(1.5,1,1.5,-1,-1.5,1,-1.5,-1,
                                         -0.5,1,0.5,1,-0.5,-1,0.5,-1,-1,0,1,0,0,0),
                                    11, 2, byrow = TRUE)
nonLinear_trajectory <- trajectory(non_linear,
                                        y0         = state,
                                        tlim       = c(0, 10),
                                        parameters = NULL,add=TRUE)
```

Observe that steady-states occur where nullclines intersect. 

Let's look at the solution curves $x(t)$ and $y(t)$: 

```{r nonlinear_pp_d}
state <- c(x=-0.5,y=1)
times <- seq(0,10,by=0.05)
non_linear_sol <- ode(y=state,times=times,func=non_linear,parms=NULL)
plot(non_linear_sol,col="blue")
```


You can learn more about the functionality of `phaseR` by reading [this page](https://www.rdocumentation.org/packages/phaseR/versions/2.1.3) or [this article](https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.660.3048&rep=rep1&type=pdf). 

# Conclusion

We have very briefly touched on the basics of R that is useful in *Topics in Biomathematics*. If there are specific topics that interest you or there is a specific type of problem you want to try to use R to solve, we recommend searching the web as there is a plethora of tutorials, documentation, and helpful discussions to be found online.  
