---
title: "R for Biomathematics"
description: |
  An Introduction to R for Biomathematics
author:
  - name: Jason M. Graham 
    url: https://topicsinbiomath.netlify.app/
    affiliation: University of Scranton 
    affiliation_url: https://sites.google.com/site/jasonmgrahamus/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 2
creative_commons: CC BY-NC
bibliography: topics_in_biomath.bib
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggformula)
library(deSolve)
library(phaseR)
```

# Computation in Biomathematics

As in other scientific disciplines, numerical and scientific computing play a central role in 
biomathematics. For the purposes of *Topics in Biomathematics*, we are mainly interested in the numerical solution of mathematical problems, *e.g.,* ordinary differential equations, and scientific visualization, *e.g.,* plotting functions. Numerous computing platforms or environments are used in biomathematics. Some notable options include

* [MATLAB](https://www.mathworks.com/products/matlab.html) or [Octave](https://www.gnu.org/software/octave/index), 

* [Maple](https://www.maplesoft.com/), 

* [Python](https://www.python.org/), 

* [XPPAUT](http://www.math.pitt.edu/~bard/xpp/xpp.html), 

* [R](https://www.r-project.org/), and

* [Julia](https://julialang.org/). 


The purpose of this document is to provide an introduction to parts of the R computing ecosystem that are particularly useful for biomathematics. 

# What is R? 

[R](https://en.wikipedia.org/wiki/R_(programming_language)) is an open-source statistical computing platform that was developed as an alternative to proprietary statistical software. In the years since it's initial release, R has become immensely popular and a vast ecosystem of open-source software has developed around R that greatly enhances it's utility in science.   

# Why R?


Given many other options, why do we focus on R as the computing tool of choice in *Topics in Biomathematics*? The principal reasons are:

1) R is free but also well-developed and widely-supported. 

2) Even though it was developed for statistical applications, R works well for numerical computing of the sort that is useful in biomathematics. 

3) The [RStudio](https://www.rstudio.com/) integrated development environment (IDE) facilitates not only coding in R but also presenting and sharing results so as to align nicely with the course objective "Communicate effectively in writing their work in applying the techniques of applied mathematics to problems arising in the life sciences.". 

4) Many biologists "speak" R and it is very helpful to have a common language when collaborating on interdisciplinary problems. 


# An Introduction to R

In what follows, we illustrate some R commands that will be used commonly throughout the *Topics in Biomathematics* course. This document is by no means comprehensive. The best ways to learn more about R are

1) by using the [`swirl`](https://swirlstats.com/) package, and 

2) searching the web for specific topics. 

Here is a short video about `swirl`:

```{r swirl_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/olOJfAjzdO8") %>%
  vembedr::use_align("center")
```


## Installing R and RStudio

R and RStudio are both free and easy to install. The videos below can serve as a helpful guide if you want to use them. Alternatively, you can access R and RStudio through a web browser without direct installation via [RStudio Cloud](https://rstudio.cloud/). This does require an account but is free to set up.

### Installation on Mac

Read [this article](https://macpaw.com/how-to/download-r-on-mac?campaign=cmmx_search_dsa_us_en&ci=10147083349&adgroupid=123618298718&adpos=&ck=&targetid=dsa-1280095891351&match=&gnetwork=s&creative=523598858635&placement=&placecat=&accname=cmm&gclid=CjwKCAiA55mPBhBOEiwANmzoQjy_N3sWjtX395AVPNBDvJVDsd2mKXLboj3XikILm3OcCx7edQqcVhoC5d8QAvD_BwE) or watch this video:

```{r mac_install,echo=FALSE}
vembedr::embed_url("https://youtu.be/AEebOXiMyyI") %>%
  vembedr::use_align("center")
```


### Installation on PC

Read [this article](https://medium.com/@GalarnykMichael/install-r-and-rstudio-on-windows-5f503f708027) or watch this video:

```{r pc_install,echo=FALSE}
vembedr::embed_url("https://youtu.be/GAGUDL-4aVw") %>%
  vembedr::use_align("center")
```

### RStudio Cloud

Here is an intro to RStudio Cloud:

```{r rstudio_cloud,echo=FALSE}
vembedr::embed_url("https://youtu.be/uK1Va_UWQFc") %>%
  vembedr::use_align("center")
```

## R as a Calculator

R can be used for basic arithmetic calculations:

```{r arithmetic}
(1 + 2) # addition
(6 - 4) # subtraction 
(3*5) # multiplication 
(8/3) # division
(2^3) # exponentiation
```

There are also a number of mathematical functions:

```{r math_functions}
(exp(1))
(log(1))
(log(exp(2)))
(log10(10))
(sin(pi/4))
(cos(pi/3))
```

Sometimes it is convenient to store a value as a variable. The follow code stores the value of 3 as a variable `x` and then multiplies it by 1.5:

```{r a_variable}
x <- 3
1.5*x
```

In R, the notation `<-` is the assignment operator.  

You can do arithmetic with variables just like you can with numbers. For example, 

```{r variable_arihtmetic}
x <- 3
y <- 10
(x + y)
(x*y)
(y^x)
```

You should always pay close attention to the order of operations when doing arithmetic in R. Use parentheses to make sure you compute what you mean to compute. For example, 

```{r op_order_1}
((3 - 2) * 6)
```

versus

```{r op_order_2}
(3 - (2 * 6))
```


## Data in R

There are several common ways to represent different types of data in R. In this section we cover those that are most relevant to biomathematics. 

### Numeric Vectors

In R, we create a vector or array of values using concatenation with the operator `c`. For example, 

```{r a_vector}
(a <- c(1,5,-7))
```

creates a numeric vector with entries 1, 5, and -7, then we store the result as a variable called `a`. 

We can perform arithmetic operations on vectors. For example, 

```{r vector_arithmetic}
b <- c(2,-1,6)
(a + b) # vector addition
(a-b) # vector subtraction
(a*b) # elementwise vector multiplication 
(a^b) # elementwise vector exponentiation 
```

**Question:** In the last code block, what is happening to produce the result corresponding to `a^b`? 

We can also call math functions on vectors. For example, 

```{r vector_math_funcs}
x <- c(1,2,3)
(exp(x))
(cos(pi*x))
```

Note that these math functions are "vectorized" in that the operate elementwise on a vector. 

### Character Vectors

We can also create so-called character vectors in R. For example, 

```{r char_vect}
(A <- c("calculus","linear algebra", "differential equations"))
```

We won't make extensive use of character vectors in *Topics in Biomathematics* but it is useful to know that they exist. 

### Matrices in R

The `matrix` functions allows us to create a matrix in R. For example, 

```{r r_matrix}
(M <- matrix(c(1,2,3,4),2,2))
(N <- matrix(c(1,2,3,4,5,6),2,3))
(P <- matrix(c(1,2,3,4,5,6),3,2))
(Q <- matrix(c(1,2,3,4,5,6),2,3,byrow=TRUE))
```

### Lists in R

All of the entries in a vector or a matrix must by of the same type. For example, all numeric values or all characters. In R, lists are a data structure that allow us to store data of mixed types. For example, 

```{r r_lists}
(l1 <- list(c(1,2,3),c("jason","graham"),matrix(c(1,2,3,4),2,2)))
(l2 <- list(initial=c(1,12),times=seq(0,10,by=1),values=matrix(1:22,11,2))) # list with names
```

Note that the entries of a list can be accessed using either indexing or names (provided that the list entries are named). For example,

```{r list_accessing}
(l1[[2]])
(l2[[1]])
(l2$initial)
```

If a list has named elements, we can access the names using the `names` function. For example, 

```{r list_names}
names(l2)
```

Here is a helpful video on R lists:

```{r list_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/Px9VNWHja4M") %>%
  vembedr::use_align("center")
```

### Data Frames in R

A data frame in R is a data structure that corresponds to something like a spreadsheet. A data frame is a special type of list object. You can think of a data frame as a cross between a list and a matrix. Here is how you ceate a data frame:

```{r data_frame}
(my_data <- data.frame(course=c("MATH 114","MATH 221","MATH 222","MATH 351", "MATH 341"),
                       num_students=c(10,12,8,5,7),
                       level=c("F","F","S","S","T")))
```

Note that a data frame contains rows and columns, every column is a vector (either numeric or character), and every column has the same number of entries. The columns of a data frame all have names. We can also use the `names` function to determine the column names of a data frame. For example, 

```{r data_frame_names}
names(my_data)
```

Typically we do not create a data frame by hand, but data is often read in as a data frame. The principal use of data frames in R is that they are the best way to provide data to plotting functions. 

Here is a helpful video on R data frames:

```{r df_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/9f2g7RN5N0I") %>%
  vembedr::use_align("center")
```

## Linear Algebra in R

Linear algebra plays a central role in solving problem in biomathematics. This section demonstrates some linear algebra computations that will be used frequently in *Topics in Biomathematics* performed in R. 

### Matrix Arithmetic

If $A$ and $B$ are matrices of the same size then we can add them. Matrix addition is computed elementwise. We demonstrate this in R.

```{r r_matrix_add}
(A <- matrix(c(1,2,-1,3,1,0,2,-1,2),3,3,byrow = TRUE))
(B <- matrix(c(3,2,-1,2,1,-1,2,-3,1),3,3,byrow = TRUE))
A + B # matrix addition
```


We can also multiply a matrix by a number (scalar multiplication). In R

```{r r_scalar_mult}
3 * A
```

Recall that if $A$ is an $m\times n$ matrix and $B$ is an $n \times p$ matrix, then we can form the [matrix product](https://en.wikipedia.org/wiki/Matrix_multiplication) $AB$ which will be a matrix of size $m\times p$. We demonstrate this in R.

```{r r_matrix_mult}
(A <- matrix(c(1,2,-1,3,1,0,2,-1,2),3,3,byrow = TRUE))
(B <- matrix(c(3,2,-1,2,1,-1),3,2,byrow = TRUE))
A %*% B # matrix multiplication
```

A special case of matrix multiplication is matrix-vector multiplication. For example, 

```{r mat_vect_mult}
(v <- c(1,-1,1))
A %*% v # matrix-vector multiplication 
```

### Matrix Transpose and Determinant

In R, the transpose of a matrix is obtained as follows:

```{r r_transpose}
t(B)
```

and the determinant of a square matrix is computed as

```{r r_det}
det(A)
```


### Eigenvalues and Eigenvectors

Recall that if $A$ is a square matrix ($n\times n$), then an eigenvalue-eigenvector pair is a number $\lambda$ and a non-zero vector $v$ satisfying $Av=\lambda v$. In R, we can compute the eigenvalues and eigenvectors for a square matrix as

```{r}
(evals_evects <- eigen(A))
```

Note that the result is an R list that consists of 

1) a vector that contains the eigenvalues, and 

2) a matrix, the columns of which are the eigenvectors.

It is a fact from linear algebra that the determinant of a square matrix is the product of its eigenvalues. Let's confirm this with our example matrix:

```{r evals_det}
(det(A))
(prod(evals_evects$values))
```

Let's also confirm that $Av=\lambda v$ for the first eigenvalue-eigenvector pair we just computed. 

```{r check_eigens}
(A %*% evals_evects$vectors[ ,1])
(evals_evects$values[1] * evals_evects$vectors[ ,1])
```

## Functions in R


Here is a helpful video on writing functions in R:

```{r functions_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/p8tAQx7ijXE") %>%
  vembedr::use_align("center")
```

## R Packages




## Visualization in R




Here is a helpful video on `ggplot2`:

```{r ggplot_video,echo=FALSE}
vembedr::embed_url("https://youtu.be/iOosTTQwCW4") %>%
  vembedr::use_align("center")
```

[This article](https://stat2labs.sites.grinnell.edu/Handouts/rtutorials/IntroToggformula.html) provides more information on `ggformula`. 

## Numerical Solution of Differential Equations in R




## Phase-Plane Methods in R






